---
title: "Study the ToothGrowth Data in R"
author: "Gurmeet Singh"
date: "November 16, 2015"
output: pdf_document
---
  
## Overview
The report investigates the dataset ToothGrowth in R and uses confidence intervals and hypothesis tests to compare affect on tooth growth by dose and type of supplement. The documentation of the ToothGrowth dataset has this to say:
```
The Effect of Vitamin C on Tooth Growth in Guinea Pigs  

Description: The response is the length of odontoblasts (cells responsible for tooth growth) in 60 
guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by 
one of two delivery methods, (orange juice or ascorbic acid (a form of vitamin C and coded as VC).

Format: A data frame with 60 observations on 3 variables.
[,1]	len	numeric	Tooth length
[,2]	supp	factor	Supplement type (VC or OJ).
[,3]	dose	numeric	Dose in milligrams/day
```
## Data Exploration
Let's plot the data using ggplot2:

```{r ref.label="ggplot_toothgrowth", echo = FALSE}
```

A graph is plotted showing the effect of vitamin C supplements on the growth of cells responsible for tooth growth. Without proceeding to any numerical analysis, conclusions can be drawn from it:

1.  Both(VC,OJ) supplements affect the growth of cells. There is a positive correlation between dosage of each and the growth of cells.
1.  At lower doses (0.5,1.0 mg/day), OJ seems to be more effective in increasing growth.
1.  At 2.0 mg/day dosage level, it is unclear whether OJ or VC is more effective but clearly VC data has higher variance compared to the OJ data.

## Statistical Analysis

### Test 1

First, let's look at whether vitamin C in higher dosages causes increase in growth (compared to a lower dose) of tooth cells by using output of the T-test. Here, the null hypothesis H0: There's no effect of higher dosage.

```
t.test(dVC2p0,dVC0p5, var.equal = FALSE)
```
Here, 

dVC2p0 = odontoblast cell lengths of guinea pigs given higher dose (2.0 mg/day) of vitamin C.

dVC0p5 = odontoblast cell lenghts of guinea pigs given lower dose (0.5 mg/day) of vitamin C.

```{r ref.label="conf_vc", echo = FALSE}
```

The confidence interval {`r conf.int.vc.vc`} is clearly above zero and shows that the mean cell length is much greater with vitamin C in dosage of 2.0 mg/day compared to 0.5mg/day. The p value (rounded to 3 digits) of `r p.vc.vc` of shows that the hypothesis H0 has very low probability of being correct, so it is rejected.

I've used ```var.equal = FALSE``` since there's nothing to suggest that the variance of the different data are equal since they may have been affected by the supplements differently.

### Test 2

Now let's compare the effect of OJ and VC on tooth growth, again using output of this T-test. Here, the null hypothesis H0: There's no effect of using vitamin C as a supplement instead of Orange Juice.
```
t.test(dOJ2p0,dVC2p0, var.equal = FALSE)
```
Where dOJ2p0 and dVC2p0 are respectively the cell lengths of guinea pigs given 2.0 mg/day of OJ and vitamin C respectively.

The T-test output is:
```{r ref.label="conf_vc_oj", echo = FALSE}
```

The confidence interval {`r conf.int.vc.oj`} contains zero and is nearly symmetric around it. This shows that there's less than 95% confidence that mean cell lengths of either set are significantly different. The p value (rounded to 3 digits) of `r p.vc.oj` of shows that the hypothesis H0 has high probability of being correct, so it is accepted, meaning that the effect of vitamin C is no more than that of orange juice.

### Test 3

Similarly, we can make vitamin C vs Orange Juice comparison at dosage level of 0.5.
The null hypothesis H0: mean of cell lengths with vitamin C dose of 0.5 mg/day is no larger than mean of cell lengths with orange juice.
```{r ref.label="conf_vc_oj_0p5", echo = FALSE}
```

The confidence interval here is {`r conf.int.vc.oj.0p5`} and does not contain zero. This shows that there's more than 95% confidence that mean cell lengths of either set are higher than the other. The p value (rounded to 3 digits) of `r p.vc.oj.0p5` shows that there's a low probability of null hypothesis being true and therefore vitamin C is more effective at increasing the cell legths than orange juice at dosage level 0f 0.5 mg/day.

All this is in agreement with what we saw earlier in the exploratory graph.

## Conclusions

1.  Increasing the vitamin C  supplement dosage has statistically significant positive effect on the odontoblast cell lengths.
1.  At lower dosages, orange juice is more effective, while at higher dosage of 2.0mg/day, vitamin C and orange juice have statistically similar effect on odontoblast cell lengths.

\pagebreak

## Appendix

The content of this report is authored in RStudio using R Markdown format and converted to PDF format using the **knitr** package. The R Markdown file itself can be found on [GitHub](https://github.com/swiftgurmeet/coursera-statistical-inference-course-project/raw/master/StatisticalInferenceCourseProject2.Rmd)

Plotting code

```{r ggplot_toothgrowth, eval = FALSE}
data(ToothGrowth)
library(ggplot2)

dVC0p5 <- ToothGrowth$len[ToothGrowth$supp == "VC" & ToothGrowth$dose == 0.5]
dVC1p0 <- ToothGrowth$len[ToothGrowth$supp == "VC" & ToothGrowth$dose == 1.0]
dVC2p0 <- ToothGrowth$len[ToothGrowth$supp == "VC" & ToothGrowth$dose == 2.0]
dOJ0p5 <- ToothGrowth$len[ToothGrowth$supp == "OJ" & ToothGrowth$dose == 0.5]
dOJ1p0 <- ToothGrowth$len[ToothGrowth$supp == "OJ" & ToothGrowth$dose == 1.0]
dOJ2p0 <- ToothGrowth$len[ToothGrowth$supp == "OJ" & ToothGrowth$dose == 2.0]

ggplot(data=ToothGrowth, aes(dose,len),colours=supp) + geom_point(col=4,pch=21) + facet_grid(. ~ supp)
```

Code for the T-tests and results' variables.

```{r conf_vc_oj, eval = FALSE}
t.test(dOJ2p0,dVC2p0, var.equal = FALSE)
conf.int.vc.oj <- t.test(dOJ2p0,dVC2p0, var.equal = FALSE)$conf.int
p.vc.oj <- round(t.test(dOJ2p0,dVC2p0, var.equal = FALSE)$p.value,3)
```

```{r conf_vc_oj_0p5, eval = FALSE}
conf.int.vc.oj.0p5 <- t.test(dOJ0p5,dVC0p5, var.equal = FALSE)$conf.int
p.vc.oj.0p5 <- round(t.test(dOJ0p5,dVC0p5, var.equal = FALSE)$p.value,3)
```

```{r conf_vc, eval = FALSE}
t.test(dVC2p0,dVC0p5, var.equal = FALSE)
conf.int.vc.vc <- t.test(dVC2p0,dVC0p5, var.equal = FALSE)$conf.int
p.vc.vc <- round(t.test(dVC2p0,dVC0p5, var.equal = FALSE)$p.value,3)
```
