---
title: "Central limit theorem simulation using exponentially distributed samples"
author: "Gurmeet Singh"
date: "November 16, 2015"
output: html_document
---
  
### Simulation to investigate the exponential distribution in R and compare it with the Central Limit Theorem

Here's an example of an exponential distribution simulated in R with n = 1000 and lambda = 0.2.
  
  
  
```{r, echo=FALSE}
expdist <- rexp(1000, 0.2)
hist(expdist,
     breaks = 20,
     main = "Exponential distribution rexp(n=1000,lambda=0.2)")
n <- 10109
```

  

The following graph is result from simulation of taking means of samples of 10000 exponential distributions each with a size of 40 and lambda = 0.2.
  
  
```{r, echo=FALSE}
n <- 40
k <- 10000
# This is defined per the project requirements
lambda <- 0.2
samplemeans <- c()
for (i in 1:k) {
  expmean <- mean(rexp(n, lambda))
  samplemeans <- c(samplemeans, expmean)
}
hist(samplemeans,breaks = 40)
```
  
  
You can see that the mean of the sample mean and the variance of the sample mean are respectively a good approximation of values predicted by the central limit theorem.
  
  

```{r, echo=FALSE}
meanofmeans <- mean(samplemeans)
sdofmeans <- sd(samplemeans)
cat("Mean of the n sample means =", meanofmeans ,"\n")
cat("Expected Mean of the n sample means (1/lambda) =", 1/lambda ,"\n")
cat("Standard Deviation of the n sample means =", sdofmeans,"\n")
cat("Expected Standard Deviation of the n sample means (1/lambda)/sqrt(n) = ", (1/lambda)/sqrt(n),"\n")
```
